---
title: "EDLD 653 Final Exam"
author: "Febe Lin"
date: today
format:
  html:
    embed-resources: true
    toc: true
execute:
  cache: true
  warning: false
  message: false
editor: visual
---

```{r}
# Load required packages.
library(httr)
library(tidyverse)
library(jsonlite)
library(lubridate)
library(tools)
library(lme4) 
library(knitr)

```

# (1) API calls to chess.com (10 points) {#api-calls-to-chess.com-10-points}

```{r}

response <- GET('https://api.chess.com/pub/country/US/players')

player_list <- fromJSON(rawToChar(response$content))$players

players_500 <- player_list[1:500]
```

```{r}

# Inspect the data structure first, then define the functions
sample_info_raw <- GET(paste0("https://api.chess.com/pub/player/", players_500[1]))
sample_info     <- fromJSON(rawToChar(sample_info_raw$content))

sample_stats_raw <- GET(paste0("https://api.chess.com/pub/player/", players_500[1], "/stats"))
sample_stats     <- fromJSON(rawToChar(sample_stats_raw$content))

str(sample_info, max.level = 3)
str(sample_stats, max.level = 3)
```

```{r}

# Helper function: return the value if available; otherwise, return NA.
# Added with assistance from Gemini to avoid NULL values in the tibble.
if_null <- function(x, default = NA) {
  if (is.null(x) || length(x) == 0) default else x
}

# Define a function to fetch information and stats for one player
get_more_details <- function(username){
  Sys.sleep(0.3)   # polite rate -limiting
  info_raw <- GET(paste0("https://api.chess.com/pub/player/", username))
  stats_raw <- GET(paste0("https://api.chess.com/pub/player/", username, "/stats"))
  
  info <- fromJSON(rawToChar(info_raw$content))
  stats <- fromJSON(rawToChar(stats_raw$content))
  
  # make a tibble to format those info and stats
  tibble(
    username          = username,
    name              = if_null(info$name, NA_character_),
    followers         = if_null(info$followers, NA_integer_),
    location          = if_null(info$location, NA_character_),
    last_online       = as_datetime(if_null(info$last_online, NA_real_), tz = "UTC"),
    rapid_last_rating = if_null(stats$chess_rapid$last$rating, NA_integer_),
    rapid_best_rating = if_null(stats$chess_rapid$best$rating, NA_integer_),
    rapid_record_win  = if_null(stats$chess_rapid$record$win, NA_integer_),
    rapid_record_loss = if_null(stats$chess_rapid$record$loss, NA_integer_),
    rapid_record_draw = if_null(stats$chess_rapid$record$draw, NA_integer_)
  )
}
 
safe_get <- safely(get_more_details)
raw_results <- map(players_500, safe_get)
player_df <- raw_results |>
  keep(~is.null(.x$error)) |>   
  map("result") |> 
  list_rbind()                  

glimpse(player_df)  
```

# (2) Identify & Remove Duplicate Files (20 points) {#identify-remove-duplicate-files-20-points}

```{r}

q2_location <- "data/question_2"

all_files <- list.files(path = q2_location, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)

cat("Initial total files detected: ", length(all_files))

# Calculate the MD5 hash for each CSV file and identify duplicate files.

hashes <- tools::md5sum(all_files)
is_duplicate <- duplicated(hashes)

# The if-else logic was added with assistance from Gemini Fast to improve code robustness when no duplicate files are found (I might re-render several times for checking or revising)

if (any(is_duplicate)) {
  dup_files  <- all_files[is_duplicate]
  dup_hashes <- hashes[is_duplicate]  
  orig_indices <- match(dup_hashes, hashes)
  orig_files   <- all_files[orig_indices]
  summary_table <- data.frame(original_file_path = orig_files,deleted_file_path  = dup_files)
  
  for (file_to_delete in dup_files) {
    file.remove(file_to_delete)
  }} else {summary_table <- data.frame(original_file_path = character(0),deleted_file_path  = character(0))
}

knitr::kable(summary_table, caption = "Duplicate Files Cleanup Ledger")


remaining_files <- list.files(path = q2_location, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)
```

```{r}
cat("Number of duplicate files deleted: ", nrow(summary_table), "\n")
cat("Number of unique files remaining:  ", length(remaining_files), "\n")

```

# **(3) Multiple Models (10 points)** {#multiple-models-10-points}

First, recode the `season` column to a new column named `wave` so that "fall" has a value of 0, "winter" has a value of 1 and "spring" has a value of 2.

```{r}
benchmarks <- read_csv("benchmarks.csv")

benchmarks <- benchmarks |>
  mutate(wave = case_when(season == "fall" ~ 0L, season == "winter" ~ 1L, season == "spring" ~ 2L))

glimpse(benchmarks)

```

```{r}

# this chunk was debug by Gemini Fast model
slope <- function(df, outcome) {
  df_score <- df |> rename(score = all_of(outcome))
  lme_mod <- lmer(score ~ 1 + wave + (1 + wave | sid),REML  = FALSE,data  = df_score)
  vc_df <- as.data.frame(VarCorr(lme_mod)) # extract the variance and make it a clear data frame
  true_var <- vc_df |>
    filter(grp == "sid", var1 == "wave", is.na(var2)) |>
    pull(vcov) # find out the True Score Variance

  # Denominator part
  individual_slopes <- df_score |>
    split(df_score$sid) |> #separate by the sid
    map_dbl(~coef(lm(score ~ wave, data = .x))["wave"]) #using OLS model
  obs_var <- var(individual_slopes) #the observed score variance
  true_var / obs_var #doing the computing
}
```

```{r}

results_q3 <- expand_grid(grade = c(3L, 4L, 5L),
  subject = c("reading", "math")) |>
  mutate(data = map(grade, ~filter(benchmarks, grade == .x)),slope_reliability = map2_dbl(data, subject, slope)) |>
  select(-data)

kable(results_q3,digits  = 3,col.names = c("Grade", "Subject", "Slope Reliability"), caption = "Slope reliability by grade and subject")
```

# **(4) Slope Reliability Function (10 points)** {#slope-reliability-function-10-points}

Write a function that takes an `lmer()` growth model object as its only argument and returns the slope reliability estimate that you just calculated. Make your function as general as possible; that is, it could work on other `lmer()` growth model objects.

Apply your function to the models in the previous question to show that it works, and show the results in a simple table.

```{r}
slope_reliability_from_lmer <- function(model) {
  df <- model.frame(model)
  outcome_var <- names(df)[1]
  grp_var <- names(getME(model, "cnms"))[1]
  slope_term  <- getME(model, "cnms")[[grp_var]][2]   
  # Numerator
  vc_df    <- as.data.frame(VarCorr(model))
  true_var <- vc_df |>
    filter(grp == grp_var, var1 == slope_term, is.na(var2)) |>
    pull(vcov)

  # Denominator
  formula_ols  <- as.formula(paste(outcome_var, "~", slope_term))
  indiv_slopes <- df |>
    split(df[[grp_var]]) |>
    keep(~nrow(.x) >= 2) |>
    map_dbl(~coef(lm(formula_ols, data = .x))[slope_term])
  obs_var <- var(indiv_slopes)
  true_var / obs_var
}
```

```{r}
# Refit the six LME models (need the model objects, not just the numbers)
models_q4 <- expand_grid(grade   = c(3L, 4L, 5L),subject = c("reading", "math")) |>
  mutate(data  = map(grade, ~filter(benchmarks, grade == .x)),
    data  = map2(data, subject, ~rename(.x, score = all_of(.y))),
    model = map(data, ~lmer(score ~ 1 + wave + (1 + wave | sid),REML = FALSE, data = .x)))

# Apply the new general function and compare with Q3 results
results_q4 <- models_q4 |>
  mutate(slope_reliability = map_dbl(model, slope_reliability_from_lmer)) |>
  select(grade, subject, slope_reliability)

kable(results_q4,digits  = 3,col.names = c("Grade", "Subject", "Slope Reliability"),caption = "Slope reliability via slope_reliability_from_lmer()")
```

# **(5) 95% Confidence Intervals via Bootstrapping (20 pts)** {#question-5-95-confidence-intervals-via-bootstrapping-20-pts}

```{r}
# Define the bootstrap function.
# The main structure of this function was developed with assistance from the Gemini Fast model.
bootstrap_data_simple <- function(df, group_var, n_samples = 100) {
  unique_ids <- unique(df[[group_var]])
  n_students <- length(unique_ids)
  student_list <- split(df, df[[group_var]])
  
  samples_list <- vector("list", n_samples)
  for (i in 1:n_samples) {
    # Keep the same random draw sequence as the original sample() call
    sampled_ids <- sample(unique_ids, size = n_students, replace = TRUE)
    sampled_dfs <- student_list[as.character(sampled_ids)]
    names(sampled_dfs) <- seq_len(n_students)
    samples_list[[i]] <- bind_rows(sampled_dfs, .id = group_var)
  }
  return(samples_list)
}
```

```{r}

slope_reliability_simple <- function(df, group_var, time_var, outcome_var) {
  
  formula_lme <- as.formula(paste(outcome_var, "~ 1 +", time_var, "+ (1 +", time_var, "|", group_var, ")"))
  
  lme_mod <- suppressMessages(suppressWarnings(lmer(formula_lme, REML = FALSE, data = df)))
  vc_df <- as.data.frame(VarCorr(lme_mod))
  
  # Debugged with assistance from Gemini.
  # The error was fixed by using vc_df$var1 instead of vc_df$name.
  true_var <- vc_df$vcov[vc_df$grp == group_var & vc_df$var1 == time_var & is.na(vc_df$var2)]
  if (length(true_var) == 0) true_var <- 0
  
  df_filtered <- df %>%
    filter(!is.na(.data[[outcome_var]]), !is.na(.data[[time_var]]))
  
  slopes_df <- df_filtered %>%
    group_by(.data[[group_var]]) %>%
    filter(n() >= 2) %>% # at least have 2 scores
    summarize(
      cov_xy = cov(.data[[time_var]], .data[[outcome_var]]),
      var_x = var(.data[[time_var]]),
      slope = cov_xy / var_x,
      .groups = "drop"
    )
  
  obs_var <- var(slopes_df$slope, na.rm = TRUE)
  return(true_var / obs_var)
}
```

```{r}
library(tidyverse)
library(lme4)

final_table_data <- tibble()
grades <- c(3, 4, 5)
subjects <- c("reading", "math")

set.seed(2653) 

for (g in grades) {
  for (s in subjects) {
    sub_data <- benchmarks %>% filter(grade == g)
    n_students <- length(unique(sub_data$sid))
    boot_samples <- bootstrap_data_simple(sub_data, group_var = "sid", n_samples = 100)
    boot_reliabilities <- c() 
    # This for loop was written with assistance from Gemini
    for (b in 1:length(boot_samples)) {
      res <- tryCatch({
        slope_reliability_simple(boot_samples[[b]], "sid", "wave", s)
      }, error = function(e) {
        return(NA)
      })
      boot_reliabilities <- c(boot_reliabilities, res)
    }
    
    # Extract clean numeric quantiles to avoid vector naming issues
    ci_lower  <- as.numeric(quantile(boot_reliabilities, 0.025, na.rm = TRUE))
    ci_upper  <- as.numeric(quantile(boot_reliabilities, 0.975, na.rm = TRUE))
    median_sr <- median(boot_reliabilities, na.rm = TRUE)
    
    current_row <- tibble(
      Grade = g,
      Subject = s,
      N = n_students,
      `95% CI Lower` = ci_lower,
      `95% CI Upper` = ci_upper,
      `Median Slope Reliability` = median_sr
    )
    
    final_table_data <- bind_rows(final_table_data, current_row)
  }
}
```

```{r}
knitr::kable(
  final_table_data, 
  digits = 3, 
  caption = "Bootstrapped (n = 100) 95% confidence intervals for slope reliability by grade and subject"
)
```
